
---
title: "Exploring gene expression and growth"
author: "Mackenzie Johnson"
date: "12/2/2021"
output: html_document
---

This document is the R version of `4-explore_expression_and_growth.ipynb`, created for easier extension of the analysis and figures created by Adam.


```{r}
library(tidyverse)


```


## Defining constants and save directories

```{r}

path_to_data <- '/Users/mackenziejohnson/Documents/grad_school/wilke_lab/Growth-expression-translation/Data/processed_data'
path_to_figures <- NULL

meta_data_files <- c(
  'processed_metadata_ecoli.tsv',
  'processed_metadata_ecoli_SPARSE.tsv',
  'processed_metadata_ecoli_PROTEIN.tsv',
  'processed_metadata_scer_RNA.tsv',
  'processed_metadata_scer_PROTEIN.tsv'
)

expression_data_files

data_key <- tibble(
  set = c('ecoli_full', 'ecoli_sparse', 'ecoli_prot',
               'scer_rna', 'scer_prot'),
  meta_file = c(
    'processed_metadata_ecoli.tsv',
    'processed_metadata_ecoli_SPARSE.tsv',
    'processed_metadata_ecoli_PROTEIN.tsv',
    'processed_metadata_scer_RNA.tsv',
    'processed_metadata_scer_PROTEIN.tsv'
  )
)

# select dataset for analysis
data_tag <- "ecoli_full"
# data_tag <- "ecoli_sparse"
# data_tag <- "ecoli_prot"
# data_tag <- "scer_rna"
# data_tag <- "scer_prot"

```


## Read in the data

```{r}

if (data_tag == "ecoli_full") {
  meta_df <- read_tsv(
    file.path(path_to_data, 'processed_metadata_ecoli.tsv')
    )
} else if (data_tag == "ecoli_sparse") {
  meta_df <- read_tsv(
    file.path(path_to_data, 'processed_metadata_ecoli_SPARSE.tsv')
  )
} else if (data_tag == "ecoli_prot") {
  meta_df <- read_tsv
}


```


## Explore the relationship between conditions

```{r}

```


#### Example scatterplots of two highly correlated conditions

```{r}

```


#### Example scatterplots of two minimally correlated conditions

```{r}

```


## Gene expression vs various dispersion metrics

```{r}

```


## Explore growth rate data

```{r}

```


#### Assess individual gene correlations with the growth rate data

```{r}

```


#### Finding representative examples of genes

```{r}

```


## Expression vs growth correlation

```{r}

```





