
---
title: "Exploring gene expression and growth"
author: "Mackenzie Johnson"
date: "12/2/2021"
output: html_document
---

This document is the R version of `4-explore_expression_and_growth.ipynb`, created for easier extension of the analysis and figures created by Adam.


```{r}
library(tidyverse)
library(glue)

```


## Defining constants and save directories

```{r}

path_to_data <- '/Users/mackenziejohnson/Documents/grad_school/wilke_lab/Growth-expression-translation/Data/processed_data/'
path_to_figures <- NULL

data_key <- tibble(
  set = c(
    'ecoli_full', 
    'ecoli_sparse', 
    'ecoli_prot',
    'scer_rna', 
    'scer_prot'
  ),
  meta_file = c(
    glue({path_to_data}, 'processed_metadata_ecoli.tsv'),
    glue({path_to_data}, 'processed_metadata_ecoli_SPARSE.tsv'),
    glue({path_to_data}, 'processed_metadata_ecoli_PROTEIN.tsv'),
    glue({path_to_data}, 'processed_metadata_scer_RNA.tsv'),
    glue({path_to_data}, 'processed_metadata_scer_PROTEIN.tsv')
  ),
  expression_file = c(
    glue({path_to_data}, 'processed_expression_ecoli.tsv'),
    glue({path_to_data}, 'processed_expression_ecoli_SPARSE.tsv'),
    glue({path_to_data}, 'processed_expression_ecoli_PROTEIN.tsv'),
    glue({path_to_data}, 'processed_expression_scer_RNA.tsv'),
    glue({path_to_data}, 'processed_expression_scer_PROTEIN.tsv')
  ),
  summary_file = c(
    glue({path_to_data}, 'processed_expression_summary_ecoli.tsv'),
    glue({path_to_data}, 'processed_expression_summary_ecoli_SPARSE.tsv'),
    glue({path_to_data}, 'processed_expression_summary_ecoli_PROTEIN.tsv'),
    glue({path_to_data}, 'processed_expression_summary_scer_RNA.tsv'),
    glue({path_to_data}, 'processed_expression_summary_scer_PROTEIN.tsv')
  )
)

# select dataset for analysis
data_tag <- "ecoli_full"

```


## Read in the data

```{r}

# get relevant file names for data set of interest
data_key %>% 
  filter(set == data_tag) -> analysis_files

# read in data
meta_df <- read_tsv(analysis_files$meta_file)

exp_df <- read_tsv(analysis_files$expression_file)

exp_sum_df <- read_tsv(analysis_files$summary_file)

```


## Explore the relationship between conditions

```{r}

```


#### Example scatterplots of two highly correlated conditions

```{r}

```


#### Example scatterplots of two minimally correlated conditions

```{r}

```


## Gene expression vs various dispersion metrics

```{r}

```


## Explore growth rate data

```{r}

```


#### Assess individual gene correlations with the growth rate data

```{r}

```


#### Finding representative examples of genes

```{r}

```


## Expression vs growth correlation

```{r}

```





