
---
title: "Exploring codon usage bias and predictors"
author: "Mackenzie Johnson"
date: "12/2/2021"
output: html_document
---

This document is the R version of `5-explore_cub_predictors.ipynb`, created for easier extension of the analysis and figures created by Adam.


```{r}
library(tidyverse)

```


## Defining constants and save directories

```{r}

```


## Data import

#### Gene-specific measurements

```{r}

```

#### Meta-data related to experiments / growth conditions

```{r}

```

#### Condition-specific expression information

```{r}

```

#### Expression sumary statistics calculated from the condition-specific expression matrices

```{r}

```

#### Merge data

```{r}

```


## Find the individual conditions whose gene expression is most (and least) correlated with CUB

```{r}

```


## Determine which expression-based summary statistics are most correlated with CUB metrics

```{r}

```


## Regression analysis to assess predictors of CUB

```{r}

```


## Robustness check

```{r}

```


## Visualizing various mult-variation models for robustness

```{r}

```


## Finally, some mixed linear models

#### Linear model with a fixed slope and random intercept

```{r}

```


#### Linear model with random slope and fixed intercept

```{r}

```

#### Linear model with random effect slope AND random intercept

```{r}

```

#### Exponential with random slopes and random intercepts

```{r}

```


